Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-07-19 paper/paper.tex
Comment out Appendix A S1 example

@@

\hypertarget{example-of-s1-for-manifold-density-estimation-todo-rewrite-as-appendix-b-with-plots}{%
\subsection{\texorpdfstring{Example of \(S^1\) for manifold density estimation {[}TODO: rewrite as appendix B with plots?{]}}{Example of S\^{}1 for manifold density estimation {[}TODO: rewrite as appendix B with plots?{]}}}\label{example-of-s1-for-manifold-density-estimation-todo-rewrite-as-appendix-b-with-plots}}

Manifold: \(S^1\) with center 0 and radius \(r=1\)

Polar coordinates: \((r, \theta)\)

Uniformly distributed angle: \(\theta \sim U(0, \pi/2)\) with \(f(\theta) = 2 / \pi\) for \(\theta \in [0, \pi /2]\).

Cartesian coordinates: \((x, y)\) where \(x = \cos(\theta) \in [0, 1]\) and \(y = \sin(\theta) \in [0, 1]\)

Density of \(\theta\): \(f(\theta) = 2 / \pi\)

Density of \(x\): \(f(x) = \frac{2}{\pi \sqrt{1-x^2}}\)

The CDF of X is given by
\[
P(X \leq x) = P(\cos(\theta) \leq x) \\
= P(\theta \geq \arccos(x)) \\
= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
= 1 - \frac{2}{\pi} \arccos(x).
\]
Then the density of X is the derivative of the CDF w.r.t. \(x\) \[
f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
= \frac{2}{\pi \sqrt{1 - x^2}}.
\]

Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could
estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\)
and compare it with its true density \(f(x)\).
We use the density estimator in \textcite{Pelletier2005-vu}. For point \(p \in S^1\), the density at \(p\) is estimated by
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
\]
where \(d=1\), \(r\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).

Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\) with \(r=1\). Therefore, the geodesic distance is
\[
d_g(p, X_i) = \arccos(p) - \arccos(X_i)
\]
As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\) equals the determinant of the metric \(g\) expressed in these coordinates at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).

{[}When using normal coordinates(orthogonal basis) for tangent vector in \(T_{X_i}S^1\), the exponential map has \emph{geodesic normal coordinates}.{]}

The Riemannian metric \(g\) for \(S^1\) is given by the inner products on the tangent spaces of \(S^1\). For each \(X_i \in M\),
\[
g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
\]
is smooth in a local neighborhood \(U\) of \(M\).

Then the volume density function \(\theta_{X_i}(p)\) is the inner product of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\), which is given by
\[
\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
\]
Therefore, the estimator for \(p \in S^1\) is given by
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
\]
where the bandwidth \(r\) is selected with the plug-in bandwidth selector implemented by the R package \emph{ks}.

\hypertarget{how-to-derive-volume-density-function-theta_x_ip}{%
\subsubsection{\texorpdfstring{How to derive volume density function \(\theta_{X_i}(p)\)}{How to derive volume density function \textbackslash theta\_\{X\_i\}(p)}}\label{how-to-derive-volume-density-function-theta_x_ip}}

\hypertarget{example-of-s2-in-henry2009-om.}{%
\paragraph{\texorpdfstring{Example of \(S^2\) in \textcite{Henry2009-om}.}{Example of S\^{}2 in @Henry2009-om.}}\label{example-of-s2-in-henry2009-om.}}

Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\). Let \(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\) is the orthonormal basis of \(\mathbb{R}^3\). Consider the exponential chart \((U, \Psi)\) induced by the parametrisation \(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by
\[
\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
\]
if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where \(r=\sqrt{s^2 + t^2}\).
Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
Then we take the partial derivatives of \(\Psi^{-1}(s,t)\) with respect to the basis of the tangent space, \(s\) and \(t\), and have that
\[
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
\]
\[
B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
\]
The coefficients for \(v\) and \(w\) gives the local coordinates of two tangent vectors.
Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\).
Therefore, we obtain that the volume density on the sphere is
\[
\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
\]
for \(q \neq p, -p\), and \(\theta_p(p) = 1\).

\hypertarget{s1}{%
\paragraph{\texorpdfstring{\(S^1\)}{S\^{}1}}\label{s1}}

Similarly in \(S^1\), we could let \(\{v, p\}\) be the orthonormal basis of \(\mathbb{R}^2\). Then the exponential chart is induced by the parametrization
\[
\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
\]
and \(\Psi^{-1}(0) = p\) where \(r = s\). Again \(r = d_g(p,q)\) if \(q = \exp_p(sv)\). Then we have
\[
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
\]
So the riemannian metric is \(g_q = \cos(r)\) and further,
\[
\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
\]
where the geodesic distance is \(d_g(p,q) = \arccos(p) - \arccos(q)\), and \(\theta_p(p) = 1\).

\clearpage

>>>>

Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \eqref{eq:denestimator}.

@@
